---
title: "Exploring the NOAA storm database"
author: "Ebberth Lopes de Paula"
date: "Friday, May 22, 2015"
output: html_document
subtitle: "An Analysis Report of Economic and Health by Severe Weather Events"
---



##*Synopsis*

The basic goal of this analysis is to explore the NOAA Storm Database.
The focus will be identify which types of events have greatest consequences in economic and in population health.
For this purpose, will be used the [Storm Database](https:\\396qusza40orc.cloudfront.net\repdata%2Fdata%2FStormData.csv.bz2) from U.S. National Oceanic and Atmospheric Administration's (NOAA).

The brief conclusion is that Tornado accounts for most deaths and injuries across America, whereas flood causes the most severe economic consequences.

##*Data Processing*


```{r, warning=FALSE, cache=TRUE}
library(dplyr)
library(ggplot2)

#Load arquive
url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
arq = "repdata-data-StormData.csv.bz2"
if (!file.exists(arq)==TRUE) {download.file(url,arq)}

data <- read.csv(bzfile(arq))
data <- tbl_df(data)

# Keeping in 'data' only the necessary information to goal
data <- select(data, EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)

#Estimated amount of damage to property by observation
data <- within(data, {
    PROPDMG_CALC <- NA
    PROPDMG_CALC[PROPDMGEXP == "B"] <- PROPDMG*10^9
    PROPDMG_CALC[PROPDMGEXP == "M" | PROPDMGEXP == "m"] <- PROPDMG*10^6
    PROPDMG_CALC[PROPDMGEXP == "K"] <- PROPDMG*10^3
    PROPDMG_CALC[PROPDMGEXP == "H" | PROPDMGEXP == "h"] <- PROPDMG*10^2
    PROPDMG_CALC[is.na(PROPDMG_CALC)] <- PROPDMG
    })


#Estimated amount of damage to crops by observation
data <- within(data, {
    CROPDMG_CALC <- NA
    CROPDMG_CALC[CROPDMGEXP == "B"] <- CROPDMG*10^9
    CROPDMG_CALC[CROPDMGEXP == "M" | CROPDMGEXP == "m"] <- CROPDMG*10^6
    CROPDMG_CALC[CROPDMGEXP == "K"] <- CROPDMG*10^3
    CROPDMG_CALC[CROPDMGEXP == "H" | CROPDMGEXP == "h"] <- CROPDMG*10^2
    CROPDMG_CALC[is.na(CROPDMG_CALC)] <- CROPDMG
    })


head(data,3)

### TOP 10 Events (summary)

## by Fatality
TOP_Fatal <- data %>%
  group_by(EVTYPE) %>%
  summarise(Fatalities = sum(FATALITIES)) %>%
  filter(Fatalities > 0) %>%
  arrange(desc(Fatalities))
TOP_Fatal <- TOP_Fatal[1:10,2]

## by Injuries
TOP_Injur <- data %>%
  group_by(EVTYPE) %>%
  summarise(Injuries = sum(INJURIES)) %>%
  filter(Injuries > 0) %>%
  arrange(desc(Injuries))
TOP_Injur <- TOP_Injur[1:10,2]

## by damage in properties
TOP_DMGPROP <- data %>%
  group_by(EVTYPE) %>%
  summarise(Prop_DMG_Value = sum(PROPDMG_CALC)) %>%
  filter(Prop_DMG_Value > 0) %>%
  arrange(desc(Prop_DMG_Value))
TOP_DMGPROP <- TOP_DMGPROP[1:10,2]

## by damage in crops
TOP_DMGCROP <- data %>%
  group_by(EVTYPE) %>%
  summarise(Crop_DMG_Value = sum(CROPDMG_CALC)) %>%
  filter(Crop_DMG_Value > 0) %>%
  arrange(desc(Crop_DMG_Value))
TOP_DMGCROP <- TOP_DMGCROP[1:10,2]


```



##*Results*
```{r}
par(mfrow = c(1, 2), mar = c(32, 6, 6, 6), mgp = c(3, 1, 0), cex = 0.8)
barplot(TOP_Fatal$Fatalities, las = 3, names.arg = TOP_Fatal$EVTYPE, main = "Top 10 Highest Fatalities", 
        ylab = "number of fatalities", col = "blue")
barplot(TOP_Injur$Injuries, las = 3, names.arg = TOP_Injur$EVTYPE, main = "Top 10 Highest Injuries", 
        ylab = "number of injuries", col = "blue")
```
